Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2015-04-13 slides/esop-termination.tex
Introduction

\section{What's The Problem?}
\begin{frame}[fragile]{Motivation}
Does this program terminate?
\begin{center}
\begin{minipage}{0.4\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
 \end{lstlisting}
\end{minipage}
\end{center}

@@

\begin{frame}[fragile]{Motivation}
Which of these guys terminates?
\begin{center}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = x * 2;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = x * 3;
}
\end{lstlisting}
\end{minipage}
\end{center}

The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
\end{frame}

@@

It's still quite hard to do for the programs that people write, which use bitvectors and all kinds of nonsense.
\end{frame}

@@

\begin{frame}{Bitvector Termination}
\begin{quote}
 ``People really don't *think* about termination using bitvector arguments''
   \hspace*\fill{\small--- Anonymous Reviewer}
\end{quote}
\end{frame}

\subsection{Non-Linearity Is Hard}

@@

\begin{frame}[fragile]{Our Theory}
 We will try to build termination proofs by solving \emph{second-order constraints}.

@@

\subsection{Ranking Functions: Proofs of Termination}

@@

\subsection{Recurrence Sets: Proofs of Non-Termination}

@@

>>>>

\section{What's Our Problem?}
\begin{frame}[fragile]{Motivation}
Does this program terminate?
\begin{center}
\begin{minipage}{0.4\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
 \end{lstlisting}
\end{minipage}
\end{center}

@@

% \begin{frame}[fragile]{Motivation}
% How about these?
% \begin{center}
% \begin{minipage}{0.45\linewidth}
%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
% while (x > 0) {
%   x = x * 2;
% }
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}{0.45\linewidth}
%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
% while (x > 0) {
%   x = x * 3;
% }
% \end{lstlisting}
% \end{minipage}
% \end{center}
% 
% \pause
% 
% The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
% \end{frame}

@@

It's still quite hard to do for the programs that people write, which use bitvectors and nonlinear operators and all kinds of nonsense.

\begin{block}{Our Goal}
 We would like to come up with a fully automatic procedure for proving termination or non-termination of
 arbitrary programs that make use of bitvector and floating point arithmetic.
\end{block}

% \vspace{1.5em}
% 
% \begin{quote}
%  ``People really don't *think* about termination using bitvector arguments''
%    \hspace*\fill{\small--- Anonymous Reviewer}
%    \end{quote}
\end{frame}

@@

% 
% \begin{frame}{Bitvector Termination}
% \begin{quote}
%  ``People really don't *think* about termination using bitvector arguments''
%    \hspace*\fill{\small--- Anonymous Reviewer}
% \end{quote}
% \end{frame}

@@

\begin{frame}[fragile]{Our Solution}
 We will try to build termination proofs by solving \emph{second-order constraints}.

@@

\subsection{Proofs of Termination}

@@

\subsection{Proofs of Non-Termination}

@@

\begin{frame}[fragile]{Oh, And One More Thing}
Does this terminate?  If so, what's the ranking function?

\begin{columns}[c]
\column{.4\textwidth}
\begin{lstlisting}[basicstyle=\normalsize]
while (x > 0 && y > 0 && z > 0) {
  if (y > x) {
    y = z;
    x = nondet();
    z = x - 1;
  } else {
    z = z - 1;
    x = nondet();
    y = x - 1;
  }
}
\end{lstlisting}

\pause
\column{.45\textwidth}
$\red{R(x, y, z) = (x < y, z)}$

\end{columns}
\end{frame}
